\chapter*{Acknowledgements}
%\addcontentsline{toc}{chapter}{Acknowledgements}

As this dissertation marks the end of my studies, it is of my responsibility to thank and acknowledge all of those who shaped the path that lead me here.
\par
First, I would like to thank my supervisor Prof. Tânia Melo and my co-supervisor Prof. Ana Maria Mendonça for all the support, dedication, and patience throughout this dissertation. They motivated me to dig deeper in the topics here approached, inspiring me and forging my knowledge, while sharing their experienced and useful insights.
\par
As I thank my supervisors, I must also thank all the professors that motivated, inspired, and, most importantly, taught me along my academic path. From English to equations and programming, the knowledge transmitted by them fills the pages of this document.
\par
I am also grateful for all the friends that made part of this process. Therefore, I thank those that I met at FEUP and made the experience in this university more fun and enjoyable. The burdens shared with them will never be forgotten, as they made the ride of the last two years smoother. I also thank those that have accompanied me since high school, who showed me the corners of this university and shared their technical insights in the topics approached in this dissertation, despite sharing a different background. It has been a pleasure to watch you grow beside me.
\par
I can not write the acknowledgments without thanking my family, especially my parents and my sister Sofia, who have been supporting me all the way through the academic path that lead me to this master's. No dissertation could have been done without your unconditional support, even when you had no clue what I was writing about.
\par
Lastly, I have to thank Mariana for all the support and companionship she provided during this journey. Especially in the last weeks, where the pressure and uncertainty started to accumulate, she made sure I kept going and gracefully completed this document. Without you, this dissertation would never happened, as you continuously encouraged me to pursuit it, believing in me more than I did. For that faith, I thank you. The growth I have experienced in these last weeks, months, and years are a reflection of your motivation to keep going and chasing my dreams, pushing me to make the most out of my capabilities. Sharing this journey and helping you whenever I could were some of the best moments I have ever lived, I am very delighted to see you succeed by my side. Without a doubt, the best and most important part of this path.

\vspace{10mm}
\flushleft{David Castanho Terroso}
